Keep your Laziness in Check

### 2018-02-04 paper/main.tex

\hzh{Rewrite counter: 2}
\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
\hzh{Okasaki has 2 functions that appends a list to the reversal of another list, one of which is too strict and the other one is just the right amount of lazy (thus incremental)}
\hzh{The amortized one is hard to specify exactly, but we can introduce approximate specification and specify it approximately}
\hzh{we need to carefully avoid saying this is a tool for specifying the time complexity of functions}
\hzh{Implementing functional programs that make clever use of laziness is very easy to get wrong, and we provide a solution to make sure the programmer got it right. We almost have no way of testing it right now before StrictCheck.}
\hzh{Email Ryan Trinkle and Dan Winograd-Cort about FRP and strictness, and try to use their examples as another problem we can solve in the intro}
\hzh{Maps will mess up memoization if you're overly strict, use knapsack as an example}
\leo{I want to keep a counter on how many times we'll rewrite this intro :)}

Lazy evaluation brings tremendous power to modularization of functional
programming ~\cite{WhyFPMatters}. Haskell is one of the most popular programming
languages that gives its programmers the power of lazy evaluation.

To demonstrate the benefits of lazy evaluaation, let us consider a dynamic
programming solution of the knapsack problem in Haskell.
%
The inputs of the knapsack problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the weight and the
value of the $i$th item. The solution presents the maximum value achievable
using a subset of the items such that their combined weight is under a limit
$W$.

The typical dynamic programming approach for knapsack maintains a table $OPT[][]$,
where the entry $OPT[j][k]$ keeps the best value achievable using only items
$1\dots j$ and a weight limit of $k$. The table satisfies the following
recurrence relation: $$OPT[j][k] = \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

With typical imperative languages, we look for shared sub-problems of the
original knapsack problem, and then we design a path along the solution table
that maximizes sharing while filling out the answers.
%
However, lazy evaluation automatically finds this path if we
use \lk{Data.Map.Lazy} and setup the solution table with the recurrence
relation \hzh{We can probably make this code prettier}
\begin{inlinecode}
ws, vs :: [Int]
n,  w  :: Int

solution :: Map (Int, Int) Int
solution = fromList [ ((j, k),
                       if (j-1) < 0 || k - ws !! j < 0
                       then 0
                       else max (solution ! (j-1, k))
                                (solution ! (j-1, k - ws !! j) + vs !! j)
                      )
                    | j <- [0..2], k <- [0..50] ]
\end{inlinecode}

@@

Unfortunately, reasoning about laziness in Haskell is difficult. Programmers
often accidentally make a program too lazy or too strict, and as a result, their
programs behave differently than expected.

>>>>

\hzh{Rewrite counter: 4}
\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
\hzh{Okasaki has 2 functions that appends a list to the reversal of another list, one of which is too strict and the other one is just the right amount of lazy (thus incremental)}
\hzh{The amortized one is hard to specify exactly, but we can introduce approximate specification and specify it approximately}
\hzh{we need to carefully avoid saying this is a tool for specifying the time complexity of functions}
\hzh{Implementing functional programs that make clever use of laziness is very easy to get wrong, and we provide a solution to make sure the programmer got it right. We almost have no way of testing it right now before StrictCheck.}
\hzh{Email Ryan Trinkle and Dan Winograd-Cort about FRP and strictness, and try to use their examples as another problem we can solve in the intro}
\hzh{Maps will mess up memoization if you're overly strict, use knapsack as an example}
\leo{I want to keep a counter on how many times we'll rewrite this intro :)}

\leo{Point of first paragraph: lazy evaluation is powerful}

Lazy evaluation brings tremendous power to functional programming
~\cite{WhyFPMatters} and is actively used in various practical
applications. For instance, the big-data framework Spark uses laziness
to drive computations on demand\cn, Apple's programming language Swift
leverages it to initialize expensive resources as needed\cn, and many
functional reactive programming web frameworks rely on it\cn \leo{This
point might need more exposition about frp}. Haskell is the most
popular programming language that allows programmers to enjoy the
power of lazy evaluation by default. Let us consider a few examples
that demonstrate the power of laziness, and remind us of the
intricacies of programming with laziness in Haskell.

\leo{Example 1: Okasaki queues}

Purely functional queue is a very popular data structure in
Haskell. Programmers typically implement such queues with two
lists---a front list holding the queue elements in FIFO order, and a
back list holding the other elments in reversed order. When the front
list is empty, a dequeue operation requires a reversal of the back
list to refill the front list. These unlucky dequeue operations take
$O(n)$ time to finish, and are much slower than their imperative
counterparts.

However, Okasaki designed a very clever purely functional queue using
lazy lists that both incrementally performs the reversal of the back
list, and incrementally appends the reversal of the back list onto the
front list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
following \lk{rotate} function to incrementally reverse and append the
back list.
\begin{inlinecode}
rotate []    back acc = (head back)  : acc
rotate front back acc = (head front) : rest
  where rest = rotate (tail front) (tail back) ((head back) : acc)
\end{inlinecode}
If we computed the expression \lk{rotate front back []}, its value
will be equivalent to \lk{front ++ (reverse back)}. However, the
function \lk{rotate} is much lazier than the naive reverse and append
operation. The function \lk{reverse} has to evaluate the entire
structure of the back list, which is not incremental. On the other
hand, the function \lk{rotate} produces a list immediately, and packs
the rest of the computation into the lazy list \lk{rest}.

Since \lk{rotate} is functionally equivalent to
\lk{front ++ (reverse back)}, a programmer could have implemented it naively
and never discover the mistake through traditional testing! The two
only differ in their laziness behavior, and there is no way of
specifying such subtle differences in any testing frameworks today.

\leo{Example 2: Knapsack}

Laziness also shines in solutions to dynamic programming problems. Let
us consider a solution to the famous knapsack problem that levareges
laziness.

The inputs of the knapsack problem are two sequences of numbers $w_i$
and $v_i$ where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify
the weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.

The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

To efficiently compute this table using imperative langauges, we look
for shared sub-problems of the original problem, and design a path
along the solution table that maximizes sharing while filling out the
answers.
%
However, lazy evaluation automatically finds this path if we
use \lk{Data.Map.Lazy} and setup the solution table with the
recurrence relation \hzh{We can probably make this code prettier}
\begin{inlinecode}
ws, vs :: [Int]
n,  w  :: Int

solution :: Map (Int, Int) Int
solution = fromList [ ((j, k),
                       if (j-1) < 0 || k - ws !! j < 0
                       then 0
                       else max (solution ! (j-1, k))
                                (solution ! (j-1, k - ws !! j) + vs !! j)
                      )
                    | j <- [0..n-1], k <- [0..w] ]
\end{inlinecode}

@@

As these examples show, the power of laziness puts great
responsibility on programmers' shoulders. An incorrect implementation
can either behave functionally the same but with inferior performance,
or it can completely break down! Reasoning about lazy programs is also
notoriously hard because predicting when a lazy value gets evaluated
is difficult.
